High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-04-26 notebooks/mnist.qmd
more work on mnist
reran model with batchsize 10

```{julia}
# Search:
generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

@@

```{julia}
# Define generator:
generator = REVISEGenerator(
  opt = opt,
  λ=0.1
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(-1,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=1,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

```{julia}
# ECCCo:
λ=[0.5,0.1,0.5]
temp=0.1
η=0.01

# Generate counterfactuals using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(η),
)

ces = Dict()
for (mod_name, mod) in model_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, mod, generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )
    ces[mod_name] = ce
end
plt_order = sortperm(collect(keys(ces)))

@@

# Set up search:
factual_label = 9
x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 7
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.9
T = 100

η=1.0

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(0.01),)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

generator = GreedyGenerator(η=η)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# ECCCo:
λ=[0.1,0.1,0.1]
temp=0.1

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(0.01),
)
ce_conformal = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using ECCCo generator:
generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=η),
)
ce_conformal_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "ECCCo", "JSMA", "ECCCo-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x_factual = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x_factual,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "eccco_mnist.png"))
```

```{julia}
# Random.seed!(1234)

# Set up search:
factual_label = 8
x_factual = reshape(counterfactual_data.X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x_factual)[1]
γ = 0.5
T = 100

# Generate counterfactual using generic generator:
generator = GenericGenerator(opt=Flux.Optimise.Adam(),)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)

# ECCCo:
λ=[0.0,1.0]
temp=0.5

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(),
)
ce_conformal = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
)
ce_conformal_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Plot:
p1 = Plots.plot(
    convert2image(MNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "cce_mnist.png"))
```

```{julia}
if M.model.model isa JointEnergyModels.JointEnergyClassifier
    jem = M.model.model.jem
    n_iter = 200
    _w = 1500
    plts = []
    neach = 10
    for i in 1:10
        x = jem.sampler(jem.chain, jem.sampling_rule; niter=n_iter, n_samples=neach, y=i)
        plts_i = []
        for j in 1:size(x, 2)
            xj = x[:,j]
            xj = reshape(xj, (n_digits, n_digits))
            plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
        end
        plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
        plts = [plts..., plt]
    end
    plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1))
    display(plt)
end
```

>>>>

```{julia}
# Search:
generic_generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

@@

```{julia}
# Define generator:
revise_generator = REVISEGenerator(
  opt = opt,
  λ=0.1
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, revise_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data_weak = deepcopy(counterfactual_data)
counterfactual_data_weak.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data_weak, M, revise_generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(-1,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

### Different Models

```{julia}
# ECCCo:
λ=[0.5,0.1,0.5]
temp=0.5
η=0.01

# Generate counterfactuals using ECCCo generator:
eccco_generator = ECCCoGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(η),
)

ces = Dict()
for (mod_name, mod) in model_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, mod, eccco_generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )
    ces[mod_name] = ce
end
plt_order = sortperm(collect(keys(ces)))

@@

### Different Generators

```{julia}
# Setup:
model = model_dict["JEM Ensemble"]

# Benchmark generators:
generator_dict = Dict(
    :wachter => generic_generator,
    :revise => revise_generator,
    :greedy => greedy_generator,
    :eccco => eccco_generator,
)

ces = Dict()
for (gen_name, gen) in generator_dict
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, model, gen; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )
    ces[gen_name] = ce
end
plt_order = sortperm(collect(keys(ces)))

plts = []
for (_name,ce) in ces
    _x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plts = plts[plt_order]
plts = [p1, plts...]
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_eccco_benchmark.png"))
```
